---
title: "STATS 506 - Homework 6"
author: "Vincent Weng"
format:
  html:
    embed-resources: true
---

<h3>Stratified Bootstrapping</h3>

Part (A) - Calculate the average RF for each team in the Fielding table. Then, since we don’t have a closed form for the standard deviation of this statistic, carry out a stratified bootstrap by team to estimate it. Do this out three ways:

1. Without any parallel processing
2. Using parallel processing with the parallel package.
3. Using futures with the future package.


Loading in data
```{r}
library(DBI)
library(RSQLite)
library(tidyverse)

# Connect to the Lahman database
con <- dbConnect(RSQLite::SQLite(), "lahman_1871-2022.sqlite")

# dbListTables(con)

fielding <- dbReadTable(con, "Fielding")

# Compute RF for each player
fielding$RF <- 3 * (fielding$PO + fielding$A) / fielding$InnOuts

# Average RF by team
team_rf <- fielding %>%
  group_by(teamID) %>%
  summarize(avg_RF = mean(RF, na.rm = TRUE))

dbDisconnect(con)
```


Without any parallel processing
```{r}
bootstrap_rf <- function(data, n_bootstrap = 1000) {
  replicate(n_bootstrap, {
    sample_data <- data %>%
      group_by(teamID) %>%
      sample_frac(replace = TRUE)
    mean(sample_data$RF, na.rm = TRUE)
  })
}

boot_results <- bootstrap_rf(fielding)
sd(boot_results)  # Standard deviation of the bootstrap samples
```



Using parallel processing with the parallel package
```{r}
library(parallel)

cl <- makeCluster(detectCores() - 1)
clusterExport(cl, "fielding")

boot_results_parallel <- parLapply(cl, 1:1000, function(i) {
  sample_data <- fielding %>%
    group_by(teamID) %>%
    sample_frac(replace = TRUE)
  mean(sample_data$RF, na.rm = TRUE)
})

stopCluster(cl)

sd(unlist(boot_results_parallel))

```


Using futures with the future package
```{r}
library(future)
library(furrr)

plan(multisession)  # Use parallel processing


boot_results_future <- future_map_dbl(1:1000, ~ {
  sample_data <- fielding %>%
    group_by(teamID) %>%
    sample_frac(replace = TRUE)
  mean(sample_data$RF, na.rm = TRUE)
})


sd(boot_results_future)

```






Part (B) - Generate a table showing the estimated RF and associated standard errors from the three approaches.
```{r}


```



Part (C) - Report and discuss the performance difference between the versions.
```{r}


```


